
title: "Download Global GBIF occurrences"
author:
- Damiano Oldoni
- Peter Desmet
- Amy Davis
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we download global occurrences for Europe from [GBIF](https://www.gbif.org). We limit the search to the species selected for risk assessment analysis and modelling.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries_download_eu}
library(tidyverse)      # To do data science
library(lubridate)      # To work with dates
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(trias)          # To use functions developed for TrIAS
library(sf)             # To work with geometries
#library(leaflet)        # To make interactive maps
library(here)
```

# Define download query parameters

## Taxa

The species selected for modelling are saved in a separate file for each group e.g `pra_plants.csv`, 'pra_mammals.csv',on GitHub:


```{r load_modelling_species_file_eu}
taxa_input_file <-here("./data/external/PRA/PRA_test.csv")
taxa <- read.csv(taxa_input_file)
```

Preview file:

```{r preview_taxa_modelling}
head(taxa)
```

Retrieve the `taxonKey`s we want to use to download occurrences:

```{r get_taxonkeys_modelling}
taxon_keys <- taxa$backbone_taxonKey # Column with taxonKeys
```

## Basis of record

All types of occurrences, except `FOSSIL SPECIMEN` and `LIVING SPECIMEN`, which can have misleading location information (e.g. location of captive animal).

```{r define_basis_of_record_eu}
basis_of_record <- c(
  "OBSERVATION", 
  "HUMAN_OBSERVATION",
  "PRESERVED_SPECIMEN", 
  "UNKNOWN", 
  "MACHINE_OBSERVATION",
  "OCCURRENCE"
)
```

## Year

Occurrences with a valid year:

```{r define_year_eu}
year_begin <- 1976
year_end <-2011
```

## Geographic coordinates

Occurrences with valid geographic coordinates:

```{r define_hasCoordinate_eu}
hasCoordinate <- TRUE
```

# Download GBIF occurrences

## Trigger download

**Note**: GBIF credentials are required in the next step. 

Trigger download:

```{r trigger_gbif_download_eu}
# Reuse existing download (comment to trigger new download)
gbif_download_key <- "0168922-220831081235567"
# Trigger new download (commented by default)
 gbif_download_key <- occ_download(
   pred_in("taxonKey", taxon_keys),
   pred_in("basisOfRecord", basis_of_record),
   pred_gte("year", year_begin),
   pred_lte("year", year_end),
  pred("hasCoordinate", hasCoordinate),
   user = rstudioapi::askForPassword("GBIF username"),
   pwd = rstudioapi::askForPassword("GBIF password"),
   email = rstudioapi::askForPassword("Email address for notification")
 )
```

## Check status of download

```{r check_download_status}
rgbif::occ_download_wait(gbif_download_key)
```

Write download to list of downloads and check pending downloads:

```{r update_download_list_eu}
#If the file gbif_downloads.tsv does not exist, create and store an empty file
if(!file.exists(here::here("data", "raw", "gbif_downloads.tsv"))) {
  download_list<-data.frame(
    gbif_download_key= character(),
    input_checklist= character(),
    gbif_download_created=character(),
    gbif_download_status=character(),
    gbif_download_doi=character()
  )
  write_tsv(download_list, here::here("data", "raw", "gbif_downloads.tsv"))
  
  remove(download_list)
}

update_download_list(
  file = here::here("data", "raw", "gbif_downloads.tsv"), 
  download_to_add = gbif_download_key, 
  input_checklist = taxa_input_file
)
```

```{r retrieve_GBIF_download}
occ_download_get(gbif_download_key, path = here("./data/raw"))
```

```{r extract_GBIF_occurrence}
raw.path<- here("data/raw//")
unzip(paste0(raw.path,gbif_download_key,".zip"),exdir=paste0(raw.path,gbif_download_key))

occ<- data.table::fread(paste0(raw.path,gbif_download_key,"/occurrence.txt"),header=TRUE)

```
#split dataframe by taxon key
```{r split_GBIF_occurrence_bySpecies}
sort(unique(occ$species))
splitDFs1<-split(occ,occ$species) 
lapply(names(splitDFs1), function(x){write.csv(splitDFs1[[x]], file = paste0(raw.path,"/",x,".csv"))})
```



